---
title: ':notebook: Code compendium for "Gaps in global non-native plant trait data and how to fill them"'
output:
  github_document:
    toc: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.13940200.svg)](https://doi.org/10.5281/zenodo.13940200)
<!-- badges: end -->

## Summary

This repository holds the compendium for the manuscript entitled "Gaps in global non-native plant trait data and how to move forward".

The goal of the manuscript and the code in this compendium is to leverage global plant trait data to depict how much we know about the traits of naturalized plants around the world. This is a prerequisite to perform any large-scale analyses about naturalized plants.

For this, we leverage data from four global plant trait databases: [AusTraits](https://austraits.org/), [BIEN](https://bien.nceas.ucsb.edu/bien/), [GIFT](https://gift.uni-goettingen.de/home), and [TRY](https://www.try-db.org/). We also use the [GloNAF](https://glonaf.org) database to get lists of naturalized species globally.

## Citation

If you want to reference this workflow you can cite it through the following citation:

> Grenié M., Bruelheide H., Dawson W., Essl F., van Kleunen M., Kühn I., Kref H., Pyšek P., Weigelt P., and Winter M.. Workflow for *Gaps in global non-native plant trait data and how to fill them*. Online at https://doi.org/10.5281/zenodo.13940200

Please also cite the preprint:

> Grenié M., Bruelheide H., Dawson W., Essl F., van Kleunen M., Kühn I., Kref H., Pyšek P., Weigelt P., and Winter M.. *Gaps in global non-native plant trait data and how to fill them*. EcoEvoRxiv preprint: https://doi.org/10.32942/X2FH0T



## Workflow structure

The workflow leverages the [`targets`](https://docs.ropensci.org/targets/) package to make reproducible analyses. The main workflow file `_targets.R` sits at the root of the repository. It defines all the the objects that are generated and that can later be accessed through `targets::tar_read()` with their names. This file refers to many different functions, abstracting away the complexity of the pipeline, that are split throughout different `.R` files in the `R/` folder.

```{r targets-glimpse, echo=FALSE}
glimpse = targets::tar_glimpse(targets_only = FALSE)

htmlwidgets::saveWidget(glimpse, "inst/glimpse.html")
```

Because the workflow is quite complex, you can visualize it [here](https://raw.githack.com/Rekyt/alientraitgaps/main/inst/glimpse.html) as a network, thanks to `targets::tar_glimpse()` from its HTML version stored in `inst/glimpse.html`?

Here's a summary of the workflow:

1. Load all trait files
    1. Load AusTraits from `austraits` package
    2. Access BIEN and GIFT databases from their R packages
    3. Load TRY database from an extract of all open information
    4. Load the Austraits Plant Dictionary (APD) from the flat file from their repository
    5. Load all the correspondence tables between databases from dedicated files
2. Access the GloNAF database
3. Harmonize the taxonomy of all databases, except BIEN, through the Taxonomic Name Resolution Service (TNRS) through the dedicated `TNRS` R package
4. Make a trait concept network and simplify all trait databases through this network
5. Collate all databases
6. Count the number of available traits per species and model it
7. Visualize these numbers and models

## Installation

The repository contains the full workflow to reproduce our analyses and figures.
You can download it through GitHub with the following link: https://github.com/Rekyt/alientraitgaps/archive/main.zip

From there, unzip it, and read the section below.

## How to run the workflow

To run it run the following command `targets::tar_make()`.
The workflow comes with an `renv` lockfile detailing all its dependencies. Use `renv::restore()` to get all packages.

The workflow needs a reliable internet connection to run.
On a standard laptop with internet, it takes about **8 hours** to complete without parallel execution.

## Needed environment variables

As the GloNAF data are not yet fully open, you need to have access to the GloNAF database and put the id and password as environment variables `GLONAF_USER` and `GLONAF_PASSWORD` respectively as provided by the GloNAF. But you can access the same data through the hosted version of GloNAF v2 on Zenodo: https://doi.org/10.5281/zenodo.14696776.

Similarly, these analyses rely on using the private access of the GIFT database to gather all observations from GIFT. To reproduce it, you need to store the private GIFT API URL as the `GIFT_RESTRICTED_API`. If you don't have access to the private API, you can use the public one, which gives very similar results.


## Trait concept network

One of the output of this work is the trait concept network to merge the four used plant trait databases (AusTraits, BIEN, GIFT, and TRY), it is available through the file named `supp_table1_trait_names_full.csv` at the following relative path `inst/exdata/correspondence_tables/supp_table1_trait_names_full.csv`. You can [download it directly](https://github.com/Rekyt/alientraitgaps/raw/refs/heads/main/inst/exdata/correspondence_tables/supp_table1_trait_names_full.csv) or used the [archived version](https://doi.org/10.5281/zenodo.13940200), or check its [online version](https://rekyt.github.io/alientraitgaps/#trait-concept-network).


## Long-term data availability

For the sake of simplicity we now record the actual version of all the data sources used here:

- AusTraits v6.0.0
- BIEN v4.2.1 
- GIFT v3.1
- TRY v6.0
- GloNAF version 2.0
- AusTraits Plant Dictionary v2.0.0
- TNRS: app.version 5.3.1, database version 4.4.1, build date = 2024-01-17, code version = 5.1, api version = 5.3.2

Because this workflow relies on several online services (BIEN, GIFT, and TNRS), re-running the workflow could actually give slightly different answers as well as may be impossible if the the online services go offline, or if the version of the code used here becomes incompatible with newer versions of the services. To ensure long-term availability of the data anda analyses provided here, the generated `_targets/` folder containing all objects generated throughout the workflow is archived [on Zenodo](https://doi.org/10.5281/zenodo.13940200). To load it, download the `_targets` folder from Zenodo then copy it at the root of the `alientraitgaps` repository and you can access all of the targets through `targets::tar_read()`.

If you reuse some of the data files provided here for scientific reproducibility, be sure to check and respect all of the licenses of all of the original data sources.


## Detailed structure of the repository

Here's a breakdown of the structure of the repository:

```
.                                  # Root of the repository
├── docs                           # GitHub pages folder to host online supplements
├── inst                           # Folder containing all data and other files
│   ├── exdata                     # Folder containing all external data
│   │   ├── apd                    # Data from AusTraits Plant Dictionary
│   │   ├── austraits              # Data from AusTraits
│   │   ├── correspondence_tables  # Correspondence tables between databases
│   │   ├── glonaf                 # Data from GloNAF
│   │   ├── gpw                    # Data on population count
│   │   ├── grip                   # Data on road density
│   │   └── try                    # Data from TRY
│   ├── figures                    # Folder with all the figures of the manuscript
│   └── presentations              # Folder containing some presentations of this work
├── manuscript                     # Manuscript files
├── R                              # All the functions needed to run the pipeline
├── renv                           # Files for the renv package
└── _targets                       # Storage for the targets package
```

## Figures and online supplements

All of the main figures of the manuscript are available as PDF and png versions in the `figures/` folder.

For the ease of reading an extended version of Figure 2 (treemap of the nubmer of species with different trait sets available), is available online at the following address:  https://rekyt.github.io/alientraitgaps/, it's also available offline as the `docs/index.html` file.


## Acknowledgements

We'd like to use our space here to acknowledge all contributors of trait observations to these databases, the people who wrote the flora aggregated by GIFT, as well as all the people who contributed aggregating and building these databases. Our work wouldn't be possible without the huge long-term collective effort of the functional trait community where each actor in the chain helped building a trait commons.
