---
title: "Global trait gaps for non-native plants: online supplement"
author: "Matthias Grenié"
bibliography: refs.bib
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document is an online supplement to the following manuscript:

> Grenié M., Bruelheide H., Dawson W., Essl F., van Kleunen M., Kühn I., Kref H., Pyšek P., Weigelt P., and Winter M.. *Gaps in global non-native plant trait data and how to fill them*. EcoEvoRxiv preprint: https://doi.org/10.32942/X2FH0T

It aims to provide some interactive figures to make easier exploration of the data produced in the paper.
It also displays the trait description network that links all the trait concepts across databases and let users download the table behind the trait description.


# Extended Figure 2 : Trait availability across taxonomy

This is an extended version of figure 2, presenting the trait combination availability across the different plant families. This figure is a treemap, where each area represents the proportion of non-native species with the given trait combination. The taxonomical families are ordered by the number of non-native species in our dataset. This extended figure uses `plotly` and help you explore the dataset and export a snapshot of the plot.

The referenced trait combinations are as follow:

- LHS trait combination: leaf mass per area, plant height, seed mass. It compares the general life history strategy of species at global scale without being context-dependent [@Westoby_Leafheightseed_1998].
- Aboveground traits: these are the six traits used in the Global Spectrum of Plant Life Form and Function [@Diaz_Global_2016]. They capture the main axes of variation in aboveground plant strategies worldwide.
- Root traits: these are the four traits defined by [@Bergmann_Fungal_2020]. They position species along the root economics spectrum and contrasts species along two axes: a “conservation” gradient corresponding to the fast-slow continuum and a “collaboration” gradient corresponding to reliance on mycorrhizal fungi to extract nutrients.

```{r plotly-fig2, fig.cap = "Treemap of trait combination availabilities across taxonomic families. The size of rectangles is proportional to the number of non-native species referenced in our dataset. The different colors represent the different available trait combinations: light gray = no trait are available for these species; darker gray = at least one trait is available in our dataset; orange = the four root traits are availabe; light purple = the three LHS traits are available; light brown = both LHS and root traits are available; dark purple = the six aboveground spectrum traits are available; dark brown = both aboveground spectrum and root traits are available.", echo=FALSE}
trait_category_color_scale = tibble::tribble(
   ~trait_category,           ~color,    ~category_name,                       
   "FALSE.FALSE.FALSE.FALSE", "#f0f0f0", "No trait",                            # No trait
   "TRUE.FALSE.FALSE.FALSE" , "#d3d3d3", "At least one trait",                 # >=1 trait
   "TRUE.FALSE.FALSE.TRUE"  , "#d25601", "Root traits",                         # Root traits
   "TRUE.TRUE.FALSE.FALSE"  , "#9283ac", "LHS",                                 # LHS
   "TRUE.TRUE.FALSE.TRUE"   , "#923601", "LHS\n& Root traits",                  # LHS + Root
   "TRUE.TRUE.TRUE.FALSE"   , "#563787", "Aboveground\nspectrum",               # Aboveground
   "TRUE.TRUE.TRUE.TRUE"    , "#551601", "Aboveground spectrum\n& Root traits" # Aboveground + Root
)

taxonomy_trait_availability = readRDS("trait_combination.Rds") |>
  dplyr::mutate(final_cat = paste(family, trait_category, sep = " - ")) |>
  dplyr::inner_join(trait_category_color_scale, by = "trait_category")

family_total = taxonomy_trait_availability |>
  dplyr::group_by(family) |>
  dplyr::summarise(n_total = sum(n_species))

fig_treemap = plotly::plot_ly(
  type = "treemap",
  ids     = c("Non-native plants", family_total$family, as.character(taxonomy_trait_availability$final_cat)),
  labels  = c("Non-native plants", family_total$family, as.character(taxonomy_trait_availability$category_name)),
  parents = c("", rep("Non-native plants", nrow(family_total)), taxonomy_trait_availability$family),
  values  = c(sum(family_total$n_total), family_total$n_total, taxonomy_trait_availability$n_species),
  textinfo="label+value+percent parent",
  insidetextfont = list(size = 14),
  outsidetextfont = list(size = 16),
  branchvalues = "total",
  pathbar = list(visible = TRUE, textfont = list(size = 18, family = "sans-serif")),
  hovertemplate = '%{label}<br>Number of species: %{value}<br>Proportion: %{percentParent:.0%}',
  marker = list(colors = c(rep("#FFF", nrow(family_total) + 1), taxonomy_trait_availability$color))
)

fig_treemap
```


# Trait concept network

We here present the trait concept network we used to combine the trait datasets.
This is the table presented as Table S1. Basically this is a combination of the
trait concepts from AusTraits, BIEN, GIFT, and TRY. The trait are matched based
on the Austraits Plant traits Dictionary (APD) and manual matching. Then all the
traits concepts that in the same connected component of the network are
considered identical. Here we aren't showing the network graphically because of
performance issues for moderately large networks (~3000 nodes and ~2000 edges).
We're interested showing the collated table version which contains the following
columns:

- `component`, number of the connected component the trait name network ordered by size.
- `consolidated_name`, automatically assigned name of the component for understanding.
- `component_size`, size of the connected component in number of nodes (= trait names)
- `austraits_trait_name`, trait name provided as in AusTraits.
- `bien_trait_name`, trait name provided as in BIEN
- `gift_trait_name`, trait name provided as in GIFT corresponding to the “Trait2” level.
- `try_trait_id`, trait id as provided in TRY TraitID column

The buttons allow you to directly download the entire table.

```{r table-trait-names, echo=FALSE}

supp_table1 = read.csv("supp_table1_trait_names_full.csv")

DT::datatable(
  supp_table1, extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print')
  )
)
```


# References

